---
title: "Economics with Virus-influenced Vector Preferences and Birth Rates and Spray Treatments: Framework #2"
author: "Richard Gomulkiewicz, Sanford Eigenbrode"
date: "Dec 2020"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r, echo=F,message=F, warning=F}
## Turn scientific notation off for printing
options(scipen = 999)
## Load the deSolve library		
library(deSolve)
```

\section{Sisterson Model with Spraying}

We use a modified version of the Sisterson model for the population dynamics of susceptible $S$ and infected $I$ vectors and the proportion $D$ of diseased hosts. Vector densities are scaled in units of the ``death free'' vector carrying capacity.  All newborns are assumed to be virus-free. The term $IM$ represents newly immigrating vectors, assumed here to be 50% infectious and 50% noninfectious. This can be modified based on measurements.

\begin{subequations} \label{unconditional}
\begin{align}
\frac{dD}{dt} & = a \frac{(1-D)}{1-D+ D\rho_+}I \\
\frac{dS}{dt} & = \left[ b_- (1-D) + D b_+ \right] T(1-T) - \delta S - \lambda \frac{D\rho_-}{1-D + D\rho_-} S + IM\\
\frac{dI}{dt} & = \lambda \frac{D\rho_-}{1-D + D\rho_-} S - \delta I + IM.
\end{align}
\end{subequations}
where $T = S + I$.

The model can be run assuming vector birth rates are the same on healthy and infected host plants ($b_I=b$) and vectors have no host preference ($\rho_- = \rho_+ = 1$). These terms can be modified to reflect known or hypothesized manipulation effects on vectors by the pathogen.

Each spray treatment instantaneously reduces the densities of vectors by the factor $1-m$. 

\bigskip

\subsection{Economics}

```{r, echo=FALSE}
# per capita yield if infected at time t
# modeled as a sigmoidal curve
k <- 1.68453
alpha <- 511.15
X <- function(t,alpha=alpha,k=k){
  #k determines the maximal steepness of increase
  #alpha^(1/k) is the the abscissa corresponding to ordinate value 1/2
  (t^k)/((alpha) + (t^k))
}
```

Let $X(t)$ be the per capita yield of a plant infected at time $t$ relative to the maximum acheivable ($0\le X(t) \le 1)$.  We use a sigmoidal curve \(X(t) = t^k/(t^k + \alpha)\) which reaches half its maximum value at $t = \alpha^{1/k}$. The parameter $k > 0$ determines the maximum steepness.

```{r, echo=FALSE}
last.time <- 150 # 210 #last day of season

time <- seq(0,last.time, by = 0.1) #time points during growing season
```


The total relative yield over a season of length $L$, $Y$, is
\begin{equation} \label{yield}
Y = \int_0^L X(\tau) \frac{dD}{dt}(\tau) d\tau + 1-D(L).
\end{equation}

To compute $Y$, we solve the initial value problem 

\begin{equation} \label{yield.ivp}
\frac{dy}{dt} = (X-1) \frac{dD}{dt} = (X-1)\frac{a (1-D)I}{1-D+D\rho_+}
\end{equation}

with initial value $y(0) = 1$ simultaneously with equations (\ref{unconditional}) over the intervals $0 \le t \le L$. Then $Y = y(L)$.

```{r, echo=FALSE}
## Solve simultaneously for D, S, and I, with different birth rates

DSIYdyn.cond <- function(t, y, p){
	D <- y[1]
	S <- y[2]
	I <- y[3]
	Y <- y[4]
	with(as.list(p),{
		dD.dt <- a*(1-D)*I/(1-D+D*rho.i)
	  dS.dt <- (b*(1-D)+bi*D)*(S+I)*(1-S-I) - delta*S - lambda*D*rho.n*S/(1-D+D*rho.n) + IM
		dI.dt <-  lambda*D*rho.n*S/(1-D+D*rho.n) - delta*I + IM
	  dY.dt <- (X(t,alpha=alpha,k=k)-1)*dD.dt
 		return(list(c(dD.dt, dS.dt,dI.dt,dY.dt)))
		})
}

#initial conditions
D0 <- 0 #disease free
S0 <- 0.01 #1% of carrying capacity
I0 <- 0.01*S0 #1% of susceptible density
Y0 <- 1

#parameters

a <- 0.2	#inoculation rate (based on Buchman et al. 2011)
delta <- 0.003 #vector death rate, estimated from longevity reported in Yang et al. 2010
lambda <- 0.2 #acquisition rate

#no preference
rho.i <- 1 #1.0
rho.n <- 1 #1.0

# case b1: b = bi
b <- .1015# this value is from Yang et al. (2010), Table 5
bi <- .07#b

# immigrating psyllids
IM <- .01

parms <- c(a = a, delta = delta, lambda = lambda, rho.i = rho.i, rho.n = rho.n, b = b, bi = bi, IM = IM)

#instantaneous mortality caused by spray treatment
m <- 0.9 

#spray treatment dates
spray.days <- c(14,21,28) #(14, 21, 28)

sprayfun <- function(t,y,parms){
  with(as.list(y),{
    y[1] <- y[1]
    y[2] <- (1-m)*y[2]
    y[3] <- (1-m)*y[3]
    y[4] <- y[4]
    return(y)
    })
}

# out <- ode(c(D0=D0, S0=S0, I0=I0), time, DSIdyn.cond, parms=parms,
#           events = list(func = sprayfun, time = spray.days))
out <- ode(c(D0=D0, S0=S0, I0=I0, Y0=Y0), time, DSIYdyn.cond, parms=parms,
           events = list(func = sprayfun, time = spray.days))

#same scenario but with no spraying
# out.nospray <- ode(c(D0=D0, S0=S0, I0=I0), time, DSIdyn.cond, parms=parms)
out.nospray <- ode(c(D0=D0, S0=S0, I0=I0, Y0=Y0), time, DSIYdyn.cond, parms=parms)

#total yield
L.index <- length(out[,1]) #index value for last day of season
yield <- out[L.index,5]
yield.no.spray <- out.nospray[L.index,5]
```

**Parameter Values**\
$L = `r last.time`$ growing season length (days)\
$a = `r a`$ vector-to-host transmission rate\
$\rho_+ = `r rho.i`$ relative preference (infected)\
$\rho_{-} = `r rho.n`$  relative preference (uninfected) \
$b = `r b`$  birth rate on healthy hosts\
$b_i = `r bi`$  birth rate on diseased hosts \
$\lambda = `r lambda`$ host-to-vector transmission rate \
$\delta = `r delta`$ vector death rate \
$\alpha = `r alpha`$ determines day at half-maximal per capita yield of a diseased host \
$k = `r k`$ steepness of per capita yield curve\
$IM = `r IM`$ immigration rates of $I$ and $S$


$m = `r m`$ vector mortality caused by single spray treatment


```{r, echo=FALSE}
max.yield.per.plant <- 1 # 1.084
spray.cost <- 50
max.value <- 3080
fixed.cost <- 1635
treat.cost <- spray.cost*length(spray.days)
net <- max.value*max.yield.per.plant*yield - treat.cost - fixed.cost 
net.no.treat <- max.value*max.yield.per.plant*yield.no.spray - fixed.cost 
```

**Yields**\
Spray Day(s): `r spray.days`

Maximum achievable yield per plant: $c = `r max.yield.per.plant`$\
Total relative yield with Spraying: $Y = `r round(max.yield.per.plant*yield,4)`$ \
Total relative yield without Spraying $Y = `r round(max.yield.per.plant*yield.no.spray,4)`$ \
Yield gain from spraying: $`r round(100*(yield - yield.no.spray)/yield.no.spray, 2)`$\%

Based on current estimates:\
Fixed costs per acre: \$`r round(fixed.cost,0)`\
Total cost of treatments per acre: \$`r treat.cost`\
Net per acre with treatment: \$`r round(net,0)`\
Net per acre with no treatment: \$`r round(net.no.treat,0)`\
Economic gain per acre from treatment: \$`r round((net-net.no.treat),0)`\/acre, or `r round(100*(net - net.no.treat)/net.no.treat, 2)`\%

\subsubsection{Per capita yield of a plant infected at time $t$, $X(t)$}
```{r, echo=FALSE}
# per capita yield if infected at time t
plot(time,X(t=time,alpha=alpha,k=k),type="l",ylab="relative per capita yield",xlab="days post emergence")
```

\subsubsection{Disease and Vector dynamics}
```{r, echo=FALSE}
#Plots of Disease and Vector Dynamics

plot(time, out[,2],type = "l", ylab = "Proportion Diseased Hosts", xlab = "Day", col = 1, ylim = c(0,1))
points(time,out.nospray[,2],type = "l",col = 2)
legend("topleft",c("treated","untreated"),lty=1,lwd=2,col=1:2,bty="n",cex=0.75)

matplot(time, cbind(out[,3], out[,4],out[,3]+out[,4]), type = "l", lty=1, ylab = "Relative Density", xlab = "Day", col = 3:1, main = "Vectors with Spraying")
legend("topleft",c("virus free","infected","total"),lty=1,lwd=2,col=3:1,bty="n",cex=0.75)

matplot(time, cbind(out.nospray[,3], out.nospray[,4],out.nospray[,3]+out.nospray[,4]), type = "l", lty=1, ylab = "Relative Density", xlab = "Day", col = 3:1, main = "Vectors without Spraying")
legend("topleft",c("virus free","infected","total"),lty=1,lwd=2,col=3:1,bty="n",cex=0.75)

```

Buchman JL, Sengoda VG & Munyaneza JE (2011) Vector transmission efficiency of beribacter by $Bactericera$ $cockerelli$ (Hemiptera: Triozidae) in zebra chip potato disease: effects of psyllid life stage and inoculation access period. Journal of Economic Entomology 104: 1486-1495. doi:10.1603/ec11123

Yang X-B, Zhang Y-M, Hua L & Liu T-X (2010) Life history and life tables of $Bactericera$ $cockerelli$ (Hemiptera: Psyllidae) on potato under laboratory and field conditions in the Lower Rio Grande Valley of Texas. Journal of Economic Entomology 103: 1729-1734. doi:10.1603/ec10083.
